---
title: "ft_analysis"
author: "Kevin O'Donnell"
date: "6/27/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

#changing name of imported dataset, removing old copy
```{r}
szn_totals <- X20_21_season_totals
remove(X20_21_season_totals)
```

#cleaning dataset to only include columns we need
```{r}

```

#defining average FT% for one proportion z-test
```{r}
avg_ft_pct <- mean(szn_totals$`FT%`, na.rm = T)
tot_fta <- sum(szn_totals$FTA, na.rm = T)
tot_ftm <- sum(szn_totals$FT, na.rm = T)
true_avg <- tot_ftm/tot_fta
```

#creating functions for 2 prop, 1 prop
```{r}
#2 prop test
prop_test <- function(x,n,p) {
  z <- ((x/n - tot_ftm/tot_fta) / sqrt((((x+tot_ftm)/(n+tot_fta))*(1-((x+tot_ftm)/(n+tot_fta)))/(1/n+1/tot_fta))))
  return(z)
}

#one prop test, output z (not p-value like generic function)
one_prop_test <- function(p1, p2, n) {
  z <- (p1-p2)/sqrt(p2*(1-p2)/n)
  return(z)
}

#outputs whether statistic is different or same as original proportion (in our case average FT%)
one_prop_test_result <- function(p1, p2, n) {
  z <- (p1-p2)/sqrt(p2*(1-p2)/n)
  ifelse(abs(z)>1.96, return("dif"), return("same"))
}

#same thing as above, but more specific
#says whether it's above or below average
one_prop_both_ways <- function(p1, p2, n) {
  z <- (p1-p2)/sqrt(p2*(1-p2)/n)
  ifelse(z>1.96, return("above average"), ifelse(between(z,-1.96,1.96), return("average"), return("below average")))
}

t_test <- function(x, u, n, s) {
  t <- (x-u)/(s/sqrt(n))
  ifelse(t>1.96, return("above average"), ifelse(between(t,-1.96,1.96), return("average"), return("below average")))
}

z_score <- function(x, u, s) {
  z <- (x-u)/(s)
  ifelse(z>1.96, return("above average"), ifelse(between(z,-1.96,1.96), return("average"), return("below average")))
}
```

#vectorizing some of the 1 prop z-test functions so we can use them in mutate statements
```{r}
vectorized_one_prop_test_result <- Vectorize(one_prop_test_result)
v_one_prop_both_ways <- Vectorize(one_prop_both_ways)
v_t_test <- Vectorize(t_test)
v_z_score <- Vectorize(z_score)

```

#putting it all together: using functions within mutate statement to create new columns to define if a player is above, at, or below average
```{r}
ft_dif <- szn_totals %>%
  mutate(dif_than_avg = vectorized_one_prop_test_result(`FT%`, true_avg, FTA)) %>%
  mutate(ft_category = v_t_test(`FT%`, true_avg, FTA, 0.1407299)) %>%
  mutate(MPG = MP/G, FTPG = FT/G, Player = sub("\\\\.*", "", Player)) 

write.csv(ft_dif,"/Users/kevinodonnell/Documents/R/NBA_Analysis/ft_unfinished.csv", row.names = FALSE)

ft_sum <- ft_dif %>%
  group_by(Player) %>%
  filter(Tm!="TOT") %>%
  summarise(across(where(is.numeric),sum)) %>%
  mutate(fg_pct=FG/FGA, three_pt_pct=`3P`/`3PA`, two_pt_pct=`2P`/`2PA`, efg=(.5*three_pt_pct+FG)/FGA, ft_pct=`FT`/`FTA`, tsa=FGA+.44*FTA, ts_pct=PTS/(2*tsa), MPG=MP/G, FTPG=FTA/G) %>%
  select(-c(`FG%`,`3P%`,`2P%`,`eFG%`,`FT%`)) %>%
  mutate(ft_category = v_t_test(ft_pct, true_avg, FTA, 0.1407299))

```

# filtering for only above, below, and average (split into 3 dataframes)
```{r}
ft_good <- ft_sum %>%
  filter(ft_category=="above average") %>%
  arrange(-ft_pct)
ft_bad <- ft_sum %>%
  filter(ft_category=="below average") %>%
   arrange(ft_pct)
ft_avg <- ft_sum %>%
  filter(ft_category=="average") %>%
   arrange(-ft_pct)

hist(ft_sum$ft_pct)
sd(ft_sum$ft_pct, na.rm = T)

ft_sum %>%
  summarise(FTA, ft_pct, ln_ft = case_when(
    ft_pct == 0 ~ 0,
    is.na(ft_pct) ~ NA_real_,
    is.numeric(ft_pct) ~ log(ft_pct))) %>%
  mutate(ft_category = v_z_score(ln_ft, log(true_avg), 0.1407299), ft_cat2 = v_one_prop_both_ways(ft_pct, true_avg, FTA)) %>%
  arrange(-ft_pct)

```

#scatterplots for above vs below avg
```{r}
ft_sum %>%
  filter(ft_category!="average", FTPG>2) %>%
  ggplot(aes(x=AST, y=TOV, color=ft_category)) +
  geom_point() +
  geom_abline(slope = .56)
  #geom_smooth(method = "lm")


  
ft_sum %>%
  filter(FTPG>2, G>15) %>%
  ggplot(aes(x=FTA, y=ft_pct, color=ft_category)) +
  geom_point() +
  geom_label_repel(data = filter(ft_sum, FTPG>2, G>30 & (FTA>500|ft_pct>.92|ft_pct<.6)),aes(label=Player), point.padding = unit(0.5, "lines"), box.padding = unit(0.5, "lines"), size=3)

ft_sum %>%
  filter(FTPG>2, G>15) %>%
  ggplot(aes(x=`3PA`, y=three_pt_pct, color=ft_category)) +
  geom_point() +
  geom_label_repel(data = filter(ft_sum, FTPG>2, G>30 & three_pt_pct>.35 & `3PA`>500), aes(label=Player), point.padding = unit(0.5, "lines"), box.padding = unit(0.5, "lines"), size=3) 

#need to figure out geom_label_repel
#change player names to before /
#add all instances of player together, redo percentages manually 

cor(ft_dif$`FT%`, ft_dif$`FG%`, use = "complete.obs")
cor(ft_dif$`FT%`, ft_dif$`3P%`, use = "complete.obs")
#corrplot(ft_dif, is.corr = F)
cor_df <- cor(ft_dif[sapply(ft_dif, function(x) is.numeric(x))], use = "complete.obs")
cor_df <- as.data.frame(cor_df)
cor_df %>%
  filter(if_any(everything(), ~.x>.8))
cor_mat <- as.matrix(cor_df)
corrplot(cor_mat, is.corr = T)
cor(ft_dif[sapply(ft_dif, function(x) is.numeric(x))], use = "complete.obs")

```

# Do players with above avg FT% shoot significantly better from 3 than below average FT shooters?
```{r}
ft_sum %>%
  group_by(ft_category) %>%
  summarise(avg_three_pt_pct = mean(three_pt_pct, na.rm = T), n())

good_avg <- mean(ft_good$three_pt_pct, na.rm = T)
bad_avg <- mean(ft_bad$three_pt_pct, na.rm = T)
avg_avg <- mean(ft_avg$three_pt_pct, na.rm = T)


#above vs below
t_test(good_avg, bad_avg, 185, sd(ft_bad$three_pt_pct, na.rm = T))

#above vs avg
t_test(good_avg, avg_avg, 185, sd(ft_avg$three_pt_pct, na.rm = T))
t_test(avg_avg, good_avg, 139, sd(ft_good$three_pt_pct, na.rm = T))

#below vs avg
t_test(avg_avg, bad_avg, 139, sd(ft_bad$three_pt_pct, na.rm = T))

```

# position distribution of above, avg, and below
```{r}
#merge with new dataset which has all their positions
positions <- read_excel("/Users/kevinodonnell/Documents/R/NBA_Analysis/20_21_season_totals\ copy.xlsx") %>%
  unique() %>%
  mutate(Player = sub("\\\\.*", "", Player))

ft_sum_positions <- left_join(positions, ft_sum, by="Player")

ft_sum_positions %>%
  filter(Pos %in% c("C", "PF", "PG", "SF", "SG")) %>%
  group_by(Pos, ft_category) %>%
  mutate(cat_n = n()) %>%
  group_by(Pos) %>%
  mutate(n=n()) %>%
  summarise(ft_category, ft_cat_prop = cat_n/n) %>%
  unique() %>%
  ggplot(aes(x=ft_category, y=ft_cat_prop)) +
  geom_col() +
  geom_text(aes(label=round(ft_cat_prop, 2)), vjust=-.5, color="blue") +
  facet_wrap(~ Pos) +
  labs(title = "Which Positions Shoot Best From the Charity Stripe?", x="Free Throw Category", y="Count") +
  scale_x_discrete(labels=c("Above", "Avg", "Below", "NA")) +
  scale_y_continuous(limits = c(0,.8))
```

## Part 2: League Free Throw Trends Over The NBA's History

#importing season averages (most stats are per 100 possessions to adjust for changes in pace except percentages usually are per game)
```{r}
league_szn_avgs <- read_excel("/Users/kevinodonnell/Documents/R/NBA_Analysis/season_averages.xlsx") %>%
  mutate(Season = as.numeric((sub("\\-.*", "", Season))))
  
```

#plotting FT/FGA over time
```{r}
league_szn_avgs %>%
  ggplot(aes(x=Season, y=`FT/FGA`)) +
  geom_line() +
  labs(title = "FT/FGA Over Time")
```

#FTA per 100 poss over time
```{r}
league_szn_avgs %>%
  filter(!is.na(FTA)) %>%
  ggplot(aes(x=Season, y=FTA)) +
  geom_line() +
  labs(title = "FTA per 100 Possessions Over Time")
```

#FT% over time
```{r}
league_szn_avgs %>%
  ggplot(aes(x=Season, y=`FT%`)) +
  geom_line() +
  labs(title = "FT% Over Time")
```

#FT% vs ORtg
```{r}
league_szn_avgs %>%
  ggplot(aes(x=ORtg, y=`FT%`)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(title = "Relationship Between Offensive Rating and FT%")

cor(league_szn_avgs$`FT%`, league_szn_avgs$ORtg, use = "complete.obs")
```

#3PA vs FTA
```{r}
league_szn_avgs %>%
  ggplot(aes(x=`3PA`, y=FTA)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(title = "Relationship Between 3PA and FTA") +
  geom_label_repel(data=filter(league_szn_avgs, Season %in% c(2020, 2010, 2000, 1990, 1980)), aes(label=Season), point.padding = unit(1, "lines"), box.padding = unit(0.5, "lines"), size=4)

cor(league_szn_avgs$`3PA`, league_szn_avgs$FTA, use = "complete.obs")

cor(league_szn_avgs$`3P`, league_szn_avgs$ORtg, use = "complete.obs")
cor(league_szn_avgs$`FT`, league_szn_avgs$ORtg, use = "complete.obs")

```

#Which players last season had above average FT/FGA? (expecting Harden, Young, Giannis, Luka?, Lebron?)
```{r}
player_averages_19 <- read_excel("/Users/kevinodonnell/Documents/R/NBA_Analysis/19_20_player_avg.xlsx")

player_averages_19 <- player_averages_19 %>%
  mutate(ft_per_fga = FT/FGA, Player = sub("\\\\.*", "", Player))

player_averages_19 %>% filter(ft_per_fga > .201, G > 30, MP > 15) %>%
  arrange(-ft_per_fga)

player_averages_19 %>% filter(ft_per_fga > .201, G > 30, MP > 15) %>%
  arrange(-FTA)

player_averages_19 %>% filter(ft_per_fga > .201, G > 30, MP > 15) %>%
  ggplot(aes(x=FTA, y=FT)) +
  geom_point() +
  geom_abline(slope=.778) +
  geom_label_repel(data = filter(player_averages_19, FTA>9), aes(label=Player), point.padding = unit(0.5, "lines"), box.padding = unit(0.5, "lines"), size=3)

player_averages_19 %>% filter(ft_per_fga > .201, G > 30, MP > 15) %>%
  ggplot(aes(x=ft_per_fga, y=FTA)) +
  geom_point() +
  geom_label_repel(data = filter(player_averages_19, (FTA>8)|(FTA>2 & ft_per_fga>.45)), aes(label=Player), point.padding = unit(0.5, "lines"), box.padding = unit(0.5, "lines"), size=3)

#this is the main one (touch up)
#if draw foul rate is how often they get to the line AND how often they convert, do I need FT%
#could just plot bar chart or switch foul rate to FTA/FGA to make it more intuitive
player_averages_19 %>% filter(G > 30, MP > 15) %>%
  ggplot(aes(x=ft_per_fga, y=`FT%`, color=FTA)) +
  geom_point() +
  geom_label_repel(data = filter(player_averages_19, (`FT%`>.9 & G>30 & ft_per_fga>.2)|(`FT%`>.5 & ft_per_fga>.4 & G>30)), aes(label=Player), point.padding = unit(0.5, "lines"), box.padding = unit(0.5, "lines"), size=3) +
  geom_vline(xintercept=.201) +
  geom_hline(yintercept = .773)
  
```



